A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-10 tex/sec5-synthesis.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’
[ tex ] Clean up code in S6

Now let us define mode-correctness rigorously.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For an argument list $\biargs$,\LT{Eww...} define its set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align}
\end{definition}
This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$.
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis-checking}.

>>>>

Now let us define mode-correctness rigorously.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For an argument list $\biargs$, define its set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align}
\end{definition}
This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$.
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \Cref{lem:soundness-completeness,thm:mode-decoration} with \Cref{thm:bidirectional-type-synthesis-checking}.
